---
title: "cm007 Exercises: Practice with `dplyr`"
output: 
  html_document:
    keep_md: true
    theme: paper
    self_contained: true
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
# use suppressPackageStartupMessages() to clean up output when loading packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tsibble))
suppressPackageStartupMessages(library(DT))
```


This worksheet contains exercises aimed for practice with `dplyr`. 


1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.

```{r}
# whatever you group by in group_by, is what summarise will summarise.
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(min_life = min(lifeExp),
            country = country[lifeExp == min_life]) %>%
    arrange(min_life)

# when you make columns with summarize, they can be immediately used in the same summarise function
```

2. Calculate the growth in population since the first year on record _for each country_ by rearranging the following lines, and filling in the `FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
DT::datatable()
group_by(country) %>% 
```

```{r}
# use first to get the population for the first year on record (current order, can use order_by to specify a different order than the current one)
gapminder %>%
    group_by(country) %>%
    arrange(year) %>%
    mutate(rel_growth = pop - first(pop)) %>%
    DT::datatable()
```


3. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent, sorted by the drop, by rearranging the following lines of code. Ensure there are no `NA`'s. Instead of using `lag()`, use the convenience function provided by the `tsibble` package, `tsibble::difference()`:

```
drop_na() %>% 
ungroup() %>% 
arrange(year) %>% 
filter(inc_life_exp == min(inc_life_exp)) %>% 
gapminder %>% 
mutate(inc_life_exp = FILL_THIS_IN) %>% 
arrange(inc_life_exp) %>% 
group_by(country) %>% 
group_by(continent) %>% 
knitr::kable()
```

```{r}
gapminder %>%
    group_by(country) %>% 
    arrange(year) %>%
    mutate(inc_life_exp = difference(lifeExp)) %>% # difference() takes vector minus its lag, equivalent to lifeExp - lag(lifeExp)
    drop_na() %>% 
    ungroup() %>% 
    group_by(continent) %>% 
    filter(inc_life_exp == min(inc_life_exp)) %>% 
    arrange(inc_life_exp) %>% # arrange works INDEPENDENTLY OF GROUPS
  # have to use .group_by = TRUE to take group into account
    knitr::kable()
```



